---
title: "OCMS WorkExp: Create MIAverse TSE"
author: "Holly Roach"
format: html
editor: source
embed-resources: true
code-fold: true
execute: 
  warning: false
---

## Aim

[Miaverse](https://microbiome.github.io/) is an R/Bioconductor framework for microbiome data analysis. Documentation and example workflows can be found at the online book, [Orchestrating Microbiome Analysis with Bioconductor](https://microbiome.github.io/OMA/docs/devel/) (Lahti et al. 2021). 

The aim of this work book is to create the [`TreeSummarizedExperiment`](https://www.bioconductor.org/packages/release/bioc//vignettes/TreeSummarizedExperiment/inst/doc/Introduction_to_treeSummarizedExperiment.html) object that is utilized for data analysis downstream.

```{r load libraries}
library(tidyverse) # for data parsing and plotting
library(ggpubfigs) # for colorblind-friendly color palettes
library(ggsignif)  # for plotting p-vals from significance testing

# for using the miaverse framework
library(mia)  
library(miaViz)
```

```{r turn off scientific notation}
options(scipen=999)
```

```{r load data}
# metadata for samples
metadata <- read_csv(
  r"(data\metadata.csv)"
  )
  
# relative abundance data for samples
relab <- read_csv(
  r"(data\metaphlan4_relative_abundances.csv)",
  )
```

## Create TSE Object

Remove any samples were no taxa have been aligned.

```{r relab w/ no assigned taxa}
# find samples where no taxa have been aligned --> suggests failed seq
failed_relab <- relab |> 
  select(where(\(x) all(x == 0)))

# exclude samples where no taxa have been aligned --> suggests failed seq
relab <- relab |> 
  select(where(\(x) !all(x == 0)))

dim(failed_relab)
```

This shows that there taxa were assigned to every samples.

To create a TSE object the following objects are required:

1.  Row data (`taxa_df`):
    - DataFrame containing taxonomic information for different taxa
    - row names should correspond to taxa IDs
    - first row (column headings) provides taxa ranks (eg Kingdom, Phylum etc)
    - row names should match 1:1 with row names in the Abundance table

```{r create rowData}
# create table showing the kingdom phylum class order family genus species for
# each taxa recorded in the microbiome data
# separate base on pattern "|"  
taxa <- data.frame(
  stringr::str_split_fixed(relab$clade_name,
  stringr::fixed("|"),
  8)
)

# add column names
names(taxa) <- c(
  "kingdom", "phylum", "class", "order", "family", "genus", "species", "strain"
)

# add a taxon column
taxa$clade_name <- relab$clade_name

# replace empty string values with NA
taxa[taxa==""] <- NA

# add column to contain taxa
taxa <- taxa |> 
  mutate(
    taxa = case_when(
      !is.na(strain) ~ strain,
      !is.na(species) ~ species,
      !is.na(genus) ~ genus,
      !is.na(family) ~ family,
      !is.na(order) ~ order,
      !is.na(class) ~ class,
      !is.na(phylum) ~ phylum,
      !is.na(kingdom) ~ kingdom)
    )

# add rownames
row.names(taxa) <- taxa$taxa

# rowData needs to be in DataFrame format
taxa_df <- DataFrame(select(taxa, -clade_name, -taxa))

# view first 3 taxa
taxa_df[1:3, 1:8]
```

2.  Abundance table (`relab_mat`):
    - data matrix containing taxa abundances for each sample
    - row names should correspond to taxa IDs
    - first row (column headings) provides sample IDs

```{r create relab_mat}
# the assays slot contains the experimental data as multiple count matrices.
# the result of assays is a list of matrices.
# the matrices can contain relative abundance, or estimated count data etc

# format relab to be suitable to add to assays
relab_mat <- relab |> 
  # remove taxon  
  select(
    - clade_name
    ) |> 
  # convert to matrix
  as.matrix() 

# add taxa names to matrix
row.names(relab_mat) <- taxa$taxa
```

3.  Col data (`metadata_df`):
    - DataFrame containing metadata relating to each microbiome sample
    - row names should correspond to sample IDs
    - first row (column headings) provides metadata variables
    - row names should match 1:1 with column heading names in the Abundance table

```{r create colData}
# remove metadata for samples with no assigned taxa 
# step not applicable

# create as dataframe (rownames is depricated with tibbles) 
metadata_df <- as.data.frame(metadata) 

# add row names
row.names(metadata_df) <- metadata_df$microbiome_id


# view metadata for first 3 participants  
metadata_df[1:3, ]
```

Overall there are: 

*    682 observations (taxa) within the row data (`taxa_df`)
*    9 observations (microbiome samples) within the col data (`metadata_df`) 
*    682 observations (taxa) and 9 variables (microbiome samples) with the abundance table (`relab_mat`).

This shows that we have data for each taxa within each sample.

**Before creating TSE object, ensure that the row and column names match one-to-one between abundance table, row data, and col data**

```{r create tse}
# ensure that the row and column names match one-to-one between 
# abundance table, rowdata, and coldata

# Match rows and columns
relab_mat <- relab_mat[rownames(taxa_df), rownames(metadata_df)]

# check coldata rownames match assay colnames
all(rownames(metadata_df) == colnames(relab_mat))

# check rowdata rownames match assay rownames
all(rownames(taxa_df) == rownames(relab_mat))

# add relab matrices to assays list
assays <- SimpleList(
  taxa_relab = relab_mat
)

#### create tse
tse <- TreeSummarizedExperiment(
  assays = assays,       
  colData = metadata_df,
  rowData = taxa_df
)

# agglomerate to taxonomy data to the species level
  tse_species <- agglomerateByRank(
    tse,
    assay.type = "taxa_relAb",
    rank = "species"
  )

# view relab data for first 3 samples and taxa stored in tse object
assay(tse_species, "taxa_relab")[1:3, 1:3]

# save TSE
saveRDS(tse_species, file = "01_create_tse/tse_species.rds")
```
