---
title: "OCMS WorkExp: Community Composition"
author: "Holly Roach"
format: html
editor: source
embed-resources: true
code-fold: true
execute: 
  warning: false
---

## Aim

This notebook will inspect the community composition using functions from [Miaverse](https://microbiome.github.io/).

In particular, we will investigate community composition using the followings functions:

1.   `plotAbundanceDensity()`
2.   `plotExpression()`
3.   `plotAbundance()`
4.   `sechm()`

Information on how to run these functions can be found in the online tutorial book, [Orchestrating Microbiome Analysis with Bioconductor](https://microbiome.github.io/OMA/docs/devel/) in:

*    [Chapter 8.3.1 Abundance](https://microbiome.github.io/OMA/docs/devel/pages/quality_control.html#abundance)
*    [Chapter 12.1 Composition Barplot](https://microbiome.github.io/OMA/docs/devel/pages/composition.html#sec-composition-barplot)

Of note, you will be applying these functions to your own data, so you do not need to load any new tse objects. Therefore, in the tutorial you can ignore the following steps:

```
data("GlobalPatterns")
tse <- GlobalPatterns
```


```{r load libraries}
library(tidyverse) # for data parsing and plotting
library(ggpubfigs) # for colorblind-friendly color palettes
library(ggsignif)  # for plotting p-vals from significance testing

# for using the miaverse framework
library(mia)  
library(miaViz)
library(scater)
```

```{r turn off scientific notation}
options(scipen=999)
```

```{r load data}
# metadata for samples
metadata <- read_csv(
  r"(data\metadata.csv)"
  )
  
# relative abundance data for samples
relab <- read_csv(
  r"(data\metaphlan4_relative_abundances.csv)",
  )

# tse object containing both the metadata and relative abundance data
# this object was created using the script found in the dir 01_create_tse
tse <- readRDS(
  r"(01_create_tse\tse_species.rds)"
)
```

## Inspect Most Abundant Microbes

Following instructions in [Chapter 8.3.1 Abundance](https://microbiome.github.io/OMA/docs/devel/pages/quality_control.html#abundance) inspect which microbes are the most abundant in your data.

Of note, inspect your metadata file to find out what groups you could use to seperate your data by. To do this, you will need to change what the `colour.by` paramter is set to.

```
# this will colour the data based on sex
colour.by = "sex"
```

```{r}
# write your code in here

```

```{r}
# write your code in here

```

```{r}
# write your code in here

```

What do you think these graphs tell us?

*write your thoughts here*

## Inspect Overall Abundance

Following instructions in [Chapter 12.1 Composition Barplot](https://microbiome.github.io/OMA/docs/devel/pages/composition.html#sec-composition-barplot) create a stacked barchart to see the overall microbiome composition per sample.

Of note, in the tutorial it mentioned transforming the data into relative abundances using the function `transformAssay()`. You did not need to this step as you tse object already contains this data.

```{r}
# write your code in here

```

```{r}
# write your code in here

```

```{r}
# write your code in here

```

What do you think these graphs tell us?

*write your thoughts here*

## Extension - Community Composition Heatmape

Following instructions in [Chapter 12.2 Composition Heatmpa](https://microbiome.github.io/OMA/docs/devel/pages/composition.html#sec-composition-heatmap) create a heatmap of overall microbiome composition per sample.


```{r}
# write your code in here

```

```{r}
# write your code in here

```

```{r}
# write your code in here

```

What do you think these graphs tell us?

*write your thoughts here*